Capturing regeneration in models of forest dynamics

### 2022-11-09 manuscript/manuscriptRW.Rmd
add Harald inputs on the result section till species diversity per main species

@@

Regeneration levels, species composition and mortality between
recruitment diameters thresholds of 7 and 10 cm varied across model
simulation results and in comparison with the observations. As expected,
recruitment was overestimated across most models for both the 7 and 10
cm diameter thresholds, with the exception of SYBILA, Landis II and
aDGVM2 that estimated recruitment rates at the lower end of the
plausibility interval of the empirical data (Figure \@ref(fig:over)).
The models with highest overestimation were PICUS and TreeMig. For most
of the models, the variability across sites in recruitment levels
(visible from the IQR
in the box plots of Figure \@ref(fig:over)) is similar to or smaller
than the observed
variability for most of the models, with the exception of PICUS, where
the mean
recruitment variability is much larger.

@@

-   Models reproduce better the dominating effect of basal area than
    effect of environmental gradients where they have varying effects
    (Figure \@ref(fig:nvTrend710BA), \@ref(fig:nvTrend710BA500) and
    FigureS 1-4). In the observed data the pattern shows that with
    increasing stand basal area the recruitment decreases, this trend
    was captured by Landis II but showing different values of basal area
    for the decreasing. Other models showed different trends such as (1)
    an increase in regeneration with increasing total basal area
    (ForCEEPS, ForCEEPS(f), FORMIND and LPJ-GUESS), (2) a rather
    constant regeneration levels with changing stand basal area (ForClim
    11, xComp, LandClim and aDGVM2) or (3) an increase of recruitment
    until certain value of stand basal area and after that a decrease in
    recruitment, but with different stand basal area thresholds (4C,
    ForClim 1, PICUS, SYBILA and iLand).

-   There are 35 sites that had a diameter threshold of 10cm and
    therefore we do not know what was the recruitment for those sites at
    7cm in the observed data, however the models did simulate
    recruitment for the 7cm threshold on those sites. The Figures
    \@ref(fig:nvTrend710BA), \@ref(fig:nvTrend710BA500) and FigureS 1-4
    show the recruitment from the diameter threshold of 7cm and those
    sites are represented in the empirical data as zeros; this would
    explain why the empirical data has zero recruitment for the lowest
    climatic water balance and highest degree day values while models do
    have recruitment.

-   The mean stand basal area across samples and per bins did not have
    the same categories in across models and empirical data. The bins
    were created using the whole stand basal area range from all the
    outputs (both from empirical and simulated data for all models). For
    those stand basal area values where the empirical data or observed
    models did not have values we assign a zero value so this stand
    basal area categories would be visible in the plots, but this zeros
    are actually not that the recruitment was zero but that this stand
    basal area category was not present for that model. This would
    explain why for the lowest stand basal areas the empirical data has
    no recruitment of for example why ,

```{r nvTrend710BA, echo=FALSE, fig.cap="Mean recruitment values across samples for the 200 sites. The values have been divided in 10 bins and the red lines represent a GAM model showing the trend from the observed data.", fig.show = 'hold', out.width = '50%'}
    knitr::include_graphics(c(here::here("figures", "envTrend7Totba.png"),
                              here::here("figures", "envTrend7Totba_500.png")))
```

## Species composition in regeneration

-   Most models increase or maintain diversity and the same happens with
    the empirical sites (Figure \@ref(fig:H710))

-   In most of the models there are small differences across sites on
    the species richness of recruitment between 7 and 10 cm (Figure
    \@ref(fig:H710). However three models, PICUS, LPJ-GUESS and Landis
    II had higher recruitment species richness than the empirical. The
    model 4C also has higher recruitment than the observed data, however
    this is a special case as this model only simulated 5 species, so
    the total species richness comparison is lower for this model.

```{r H710, echo=FALSE, fig.cap="Shannon index of the recruitment calculated by basal area as the mean value across samples per site for diamter treshold of 7 and 10cm. The red dotted lines represent the 20 percentile of the observed data for the treshold of 7cm and 70 percentile of the observed data for the treshold of 10cm ", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

-   Most of the models are capturing well or over-predicting the species
    richness at the stand level (Figure \@ref(fig:combinedRichness) A
    and FigureS 8).

-   The models over predicting species richness at the stand level are
    all the "ForClim family" and FORMIND ] (FigureS 8)

-   Only few models under-predict the species richness at the stand
    level. LPJ-GUESS this is a global model that might not be design to
    capture well this, ForCEEPS (f), xComp, iLand and 4C that did
    simulate less species than the ones available (Figure
    \@ref(fig:combinedRichness), and FigureS 8).

-   Species composition is lower in the recruitment compared to the
    stand level, both for the observed data and simulated data and both
    for recruitment thresholds of 7 and 10 cm. (Figure
    \@ref(fig:combinedRichness)B, and FigureS 9)

-   Most of the models capture correctly the recruited species richness
    at lower values (Figure \@ref(fig:combinedRichness)B, and FigureS 9)

-   Many of the models underestimate species richness in recruitment
    (Figure \@ref(fig:combinedRichness)B, and FigureS 9)

-   Models simulating higher diversity at the stand do not have higher
    diversity in recruitment (FigureS 6 an 7).

-   Both models with feedback and without feedback from the adult trees
    underestimate species richness in recruitment (which ones have
    feedback! enumerate them)) (Appendix figure?)).

```{r combinedRichness, echo=FALSE, fig.cap="Mean Shannonn index across all samples per site and for observed and simulated values. A) Shows the three trends present across models for the species richness at the stand level. B) Shows the three trends present across models for recruitment at 7cm treshold. The selection representative model per trend group was based on the Figure XX and tables available in the supplementary information", out.width='75%'}
    knitr::include_graphics(here::here("figures", "combined_richness.png"))
```

@@

>>>>

There are 35 sites that had a diameter threshold of 10 cm and therefore
we do not know the actual recruitment rates for those sites at 7 cm; yet
the models simulated recruitment for the 7 cm threshold on those sites.
The 35 sites were removed from the simulated data as well to make a fair
comparison.

How many species were simulated per model? why?

Need to say something about zero inflation in the observations vs. the
simulations (current Figure S5)

@@

Regeneration levels, species composition and mortality between
recruitment diameters thresholds of 7 and 10cm varied across model
simulation results and in comparison with the observations. As expected,
recruitment was overestimated across most models for both the 7 and 10
cm diameter thresholds, with the exception of SYBILA, Landis II and
aDGVM2 that estimated recruitment rates at the lower end of the
plausibility interval of the empirical data (Figure \@ref(fig:over)).
The models with highest overestimation were PICUS and TreeMig. For most
of the models, the variability across sites in recruitment levels
(visible from the IQR in the box plots of Figure \@ref(fig:over)) is
similar to or smaller than the observed variability for most of the
models, with the exception of PICUS, where the mean recruitment
variability is much larger.

@@

When looking at the representation of recruitment rates along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (water balance), distinct features emerge: the models
reproduce the effect of basal area much better than the effects of
climatic gradients, where they feature varying effects (Figure
\@ref(fig:nvTrend710BA), \@ref(fig:nvTrend710BA500) and FigureS 1-4).

In the observed data the pattern shows that with increasing stand basal
area recruitment decreases. However, this trend was captured by Landis
II only, albeit at much lower values of basal area than in the
observations. Other models featured distinctly different trends, such as
(1) an increase in regeneration with increasing total basal area
followed of no recruitment at high basal area (ForCEEPS, ForCEEPS(f),
FORMIND, TreeMig and LPJ-GUESS), (2) almost constant regeneration levels
with basal area (ForClim 11, xComp, LandClim and aDGVM2) or (3) an
increase of recruitment up to a certain value of stand basal area
followed by a decrease of recruitment at even higher stand basal area
values at model-specific thresholds (4C, ForClim 1, PICUS, SYBILA and
iLand).

Models did not match the observed patterns in recruitment across soil
moisture and temperature gradients and there was either a clear common
pattern across models. Looking at the soil moisture gradients (FigureS
2), we can see that a group of models decreases recruitment with
increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS
and LandClim), while few models have the opposite trend (ForCEEPs(f),
FORMIND and LPJ-GUESS), and the rest of do not have a clear pattern
(ForCEEPS, iLand, Landis II, TreeMig and aDGVM2). Across temperature
gradients all the models that had a decrease in recruitment with
increasing soil moisture showed the opposite trend (more recruitment
with increasing temperature) except for 4C (FigureS 3). The rest of the
models did not have a clear patterns across the temperature gradient.

```{r nvTrend710BA, echo=FALSE, fig.cap="Mean recruitment values across samples for the 200 sites against the stand basal area. (A) y-axis scaled by model; (B) same y-axis across all models. The values were split into 10 bins; the red lines represent a GAM model showing the trend in the observed data.", fig.show = 'hold', out.width = '80%'}
    knitr::include_graphics(c(here::here("figures", "envTrend7Totba.png"),
                              here::here("figures", "envTrend7Totba_500.png")))
```

Most models match the level of diversity (as measured by the Shannon
index) of the observed data quite well (Figure \@ref(fig:H710)),
although they consistently underestimate its variability across sites.
Four models overestimate recruitment diversity (4C, PICUS, LANDIS-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case as this model simulated 5 species only. In most
models there are small differences across sites regarding the species
richness of recruitment between 7 and 10 cm. Most models maintain or
even increase in recruitment diversity between the 7 and 10 cm
thresholds, and the same is visible from the empirical data, however the
differences are not significant EXPLAIN WHY  "Can we understand this (I
know this is for the Discussion, but still...)?? This cannot be species
richness (as there are no new species popping between 7 and 10 cm DBH).
I'd also expect the number of species to be lower at 10 cm than at 7 cm
(because some will die off due to insufficient light etc.), but this is
not reflected in H' because species with zero abundance are ignored
(ln[0] is not defined). So a strongly increasing evenness must be
driving this increase of H'."

```{r H710, echo=FALSE, fig.cap="Shannon index of the recruitment, calculated by basal area as the mean value across samples per site for the diameter thresholds of 7 and 10 cm, respectively. The red dotted lines represent the 20th percentile of the observed data for the threshold of 7 cm and the 70th percentile of the observed data for the threshold of 10 cm.", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

Most of the models are capturing well or over-predicting the species
richness at the stand level (Figure \@ref(fig:combinedRichness) A2 and
FigureS 6). The models over predicting species richness at the stand
level are ForClim1, ForClim 11, FORMIND and TreeMig (Figure
\@ref(fig:combinedRichness) A1). However only few models under-predict
the species richness at the stand level (Figure
\@ref(fig:combinedRichness) A3) LPJ-GUESS, ForCEEPS (f), xComp, iLand
and 4C. Species diversity is lower in the recruitment compared to the
stand level, for both observed and simulated data, and for both
recruitment thresholds, i.e. 7 and 10 cm (Figure
\@ref(fig:combinedRichness)B, and FigureS 7). The majority of the models
capture reasonably well the recruitment diversity at lower values of
diversity, however most of them strongly underestimate species diversity
in recruitment for higher observed diversity levels. Models simulating
higher diversity at the stand level do not feature higher diversity in
recruitment
(FigureS 6 and 7). Both models with feedback and without feedback from
the adult trees (via seed production) underestimate species richness in
recruitment.

```{r combinedRichness, echo=FALSE, fig.cap="Mean Shannon index across all samples per site and for observed and simulated values. A) Shows the three trends present across models for the species richness at the stand level. B) Shows the three trends present across models for recruitment at 7cm threshold. The selected representative model per trend group was based on Figure XX and tables available in the supplementary information", out.width='75%'}
    knitr::include_graphics(here::here("figures", "combined_richness.png"))
```

@@

-   Also BTW, the reason why some models have zero regeneration at very
    low BA values may be that they never actually simulate such low BA
    values at the one ha level (which was the sampling unit if I recall
    correctly). If so, this should be pointed out as well, I think.
